### 2015-12-02 paper/3-features.tex
Some broken sentences rewritten.

\subsubsection{Optimistic}
The optimistic kernel requires from the hosted model only that copying the state is well-defined, which is provided in the base State class for the user. 
The kernels use Mattern's \cite{mattern}
GVT algorithm with a maximum of 2 rounds per iteration to determine a GVT. This process runs asynchronously from the simulation itself. Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states/(anti)messages.

@@

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks: 
\subsubsection{Heap}
A kernel never sends a complete object to another kernel, only a pointer to the object. This avoids a possibly expensive copy of the payload, but at the cost of allocation overhead.\\
This cost becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and messages, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}. In this way allocating threads do not block each other, and in a single threaded kernel we can leverage arena-style pools if desired.\\
This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (if it can prove it is no longer used). Experiments with synchronized pools proved slower than malloc/free. \\
The kernels initially used strings as identifiers, as is done in PyPDEVS. Profiling quickly indicated this to be a performance bottleneck. C++ strings are heap allocated variable sized objects with an atomic reference count. Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new object, or copy existing.\\
Strings are more intuitive to work with from a user's standpoint. So as a compromise the user can reference models/ports by string name (usually when constructing the model). Once simulation starts all objects use integral identifiers for performance. This also increased usage of the constexpr feature of C++11 in, amongst others, timestamps and message headers.
\\
\subsubsection{Raw pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads, the reference counting becomes very expensive, and the (de)allocating of memory caused significant contention between threads. Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it. Messages are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the timestamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
\subsubsection{Schedulers}
PyPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type. Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before. Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the amount of items to update. Items are only updated, never removed.

>>>>

\subsubsection{Optimistic}
The optimistic kernel requires from the hosted model only that copying the state is well-defined, which is provided in the base State class for the user. 
The kernels use Mattern's \cite{mattern}
GVT algorithm with a maximum of 2 rounds per iteration to determine a GVT. This process runs asynchronously from the simulation itself. Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states and (anti)messages.

@@

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks: 
\subsubsection{Heap}
Events are passed by pointer, avoiding a possibly expensive copy of the payload at the cost of allocation overhead.\\
Allocation becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and events, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}. In this way allocating threads do not block each other. If desired, arena-pools are available for single-threaded simulation. A disadvantage is increased complexity in ownership semantics. The creating kernel is responsible for destruction, but this can only be guaranteed up to GVT. Experiments with synchronized pools proved slower than malloc/free. \\
The kernels initially used strings as identifiers, as is done in PyPDEVS. Profiling quickly indicated this to be a performance bottleneck. C++ strings are heap allocated variable sized objects with an atomic reference count, not immutable objects as in Python. Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new objects, or copy existing.\\
Strings are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name. Once simulation starts all objects use integral identifiers for performance. This also increased usage of the constexpr feature of C++11 in, amongst others, timestamps and message headers.
\subsubsection{Raw pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads, the reference counting becomes very expensive, and the (de)allocating of memory caused significant contention between threads. Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it. Messages are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the timestamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
\subsubsection{Schedulers}
PyPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type. Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before. Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the amount of items to update. Items are only updated, never removed.
